An efficient feedback calibration algorithm for direct imaging radio telescopes

### 2016-03-02 cal_demo/ms.tex
addressing more comments, tweaking schematic figure

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients, which are used to estimate the antenna gains. These gain estimates are then fed back into the correlator to be applied to subsequent data, and the process is repeated. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip". ({\bf I have some suggestions to tweak this figure.})}
\label{fig:schematic}
\end{center}
\end{figure}

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the damped version is essentially a weighted average over the past several iterations, giving it a longer effective integration time.

@@

\section{Discussion}\label{sec:discussion}
Through simulations and application to real data, we have shown that the EPICal algorithm is a viable solution for calibrating direct imaging arrays in real time. The computation necessary only scales with the number of antenna elements, making it a sub-dominant cost factor when designing the correlator. This strategy will enable fast read-out and arrays with many thousands of antennas, which will be necessary for future radio transient and cosmology experiments respectively.

@@

\textbf{Fitting gain models.} With some knowledge of the instrumental bandpass, the signal-to-noise of the gain solutions ({\bf you mean of the image?}) can be greatly improved by fitting a model to the per-frequency solutions derived here. We used this method at a rudimentary level in our demonstration to LWA data by assuming the gains were constant over a narrow bandwidth. One could easily improve on this by extending the bandwidth and fitting for a low order polynomial in phase and amplitude. Additionally, with knowledge of the poly-phase filters applied to the data stream, we could include channels closer to the edge of the band.

@@

\textbf{Multiple pixel correlations.} As pointed out in section~\ref{sec:noise}, for a sky-noise dominated system, the signal to noise in the calibrator pixel dominates the error on the EPICal generated gain estimates. Besides the methods described above to increase signal to noise, one could also use correlations of multiple image pixels to incorporate a higher fraction of the total sky noise. Of course this increases the computational cost to a scaling of $\mathcal{O}(\Nant N_{\mathrm{pix}})$, which will typically be much lower than the $\mathcal{O}(\Ng \log_2 \Ng)$ of the correlator itself.

\textbf{Direction dependent gains.} Even below the timescale of a dynamic ionosphere, a widefield instrument may require direction dependent gains to account for a varying thickness ionospheric sheet. The holographic mapping framework used by the MOFF algorithm, and in turn the EPIC and EPICal software can naturally account for direction dependence through an updated model of the antenna voltage pattern. This can be achieved by using multiple pixel correlations to independently solve for gains in each pixel direction. These solutions can then be used to fit a beam model on the sky (analogous to the gain model fitting mentioned above), and updating the antenna voltage pattern. This updated pattern would then feed into the gridding step of the correlator, allowing the imager to convolve the signals with the effective beam pattern in the ground plane.

The work here will serve as a foundation for further development. We have shown that the EPICal algorithm produces reliable calibration solutions, and have identified several aspects to increase the scope. With next generation instruments in the planning and development stages, EPICal is poised to greatly contribute to their success.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figures/schematic.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel from the (unsquared) image is tapped out and correlated against the input antenna electric field signals to form $\Cna$ coefficients (equation~\ref{eq:Cna}). These coefficients are then shipped off-chip to apply equations~\ref{eq:cal_solution}~and~\ref{eq:damping}, and any other processing to arrive at an updated estimate of the gains, $g^{(n)}_a$. The gray box shows operations which must be done at high speed (before averaging in time), which white boxes show operations which can be performed ``off-chip". ({\bf I have some suggestions to tweak this figure.})}
\label{fig:schematic}
\end{center}
\end{figure}

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it was iterated, as is often the case in iterative minimization methods. To mitigate this we introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, effectively giving the solutions memory of previous iterations.
\begin{equation}\label{eq:damping}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is actually used in the iterative calibration loop. Once the loop converges the damped version is essentially a weighted average over the past several iterations, giving it a longer effective integration time.

@@

\section{Discussion}\label{sec:discussion}
Through simulations and application to real data, we have shown that the EPICal algorithm is a viable solution for calibrating direct imaging arrays in real time. The computation necessary only scales with the number of antenna elements, making it a sub-dominant cost factor when designing the correlator. This strategy will enable fast read-out for arrays with many thousands of antennas, which will be necessary for future radio transient and cosmology experiments.

@@

\textbf{Multiple pixel correlations.} As was seen in section~\ref{sec:noise}, the single pixel correlation demonstrated in this paper can underperform in the presence of a complex sky, when compared to visibility-based gains with perfect knowledge of the sky. These errors can be mitigated by using correlations of multiple image pixels to incorporate a higher fraction of the total sky power into the calibration loop. Of course this increases the computational cost to a scaling of $\mathcal{O}(\Nant N_{\mathrm{pix}})$, which will typically be much lower than the $\mathcal{O}(\Ng \log_2 \Ng)$ of the correlator itself. These additional correlations would further enable direction dependent gain solutions. Each correlation could be used to independently solve for gains in each pixel direction, then fit to a beam model on the sky. This updated beam pattern would then feed into the gridding step of the correlator, allowing the imager to convolve the signals with the effective beam pattern in the ground plane.

\textbf{Fitting gain models.} With some knowledge of the instrumental bandpass, the noise of the gain solutions can be greatly reduced by fitting a model to the per-frequency solutions derived here. We used this method at a rudimentary level in our demonstration to LWA data by assuming the gains were constant over a narrow bandwidth. One could easily improve on this by extending the bandwidth and fitting for a low order polynomial in phase and amplitude. Additionally, with knowledge of the poly-phase filters applied to the data stream, we could include channels closer to the edge of the band.

@@

The work here will serve as a foundation for further development. We have shown that the EPICal algorithm produces reliable calibration solutions, and have identified several aspects to increase the scope. With next generation instruments in the planning and development stages, EPICal is poised to greatly contribute to their success. The software is integrated into the EPIC package and freely available to use in simulations or offline processing of data. Work is underway to port the code to real-time GPU systems for deployment.
